---
title: "BIDS conversion example with toy data"
author: "Dr. Xaver Fuchs"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A_Example_with_toy_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(r2bids)
```


In this document, we demonstrate how to use the `create_bids` function. This function generates a BIDS-compliant dataset, including participant and task event files. We will walk through the following steps:

1. Create toy data
2. Validate the input data
3. Write the BIDS data
4. Define and write JSON metadata
5. Read and inspect the generated files

## Step 1: Create Toy Data

We will create a simple toy dataset with participant information and behavioral data.
Note that some variables are deliberately not called like they should be called according to BIDS conventions.

```{r}
# Toy dataset
example_data <- data.frame(
  ParticipantID = c(1, 2, 3, 1, 2, 3),
  Sex=c("M", "F", "O", "M", "F", "O"),
  Age=c(21, 32, 27, 21, 32, 27),
  Session = c(1, 1, 1, 2, 2, 2),
  ResponseTime = c(350, 400, 375, 415, 372, 401),
  Accuracy = c(1, 0, 1, 1, 1, 1)
)

# View the data
print(example_data)
```

The dataset contains three participants with their response times and accuracy. Each participant completed one session of a reaction time task.


## Step 2: Validate Input Data

Before creating a BIDS dataset, it's important to ensure that the input data complies with naming conventions (snake_case) and contains necessary columns like participant IDs and sessions. Some checking is implemented in the check_input_data() function that renames variables and labels to snake_case and reports that. Note that it creates a new data set. Please check the resulting set to ensure that everything is correct.

Here we will apply that function to the toy data created in the last step.

```{r}
# Run validation
example_data_checked <- check_input_data(data = example_data, 
                 participant_col = "ParticipantID", 
                 session_col = "Session", 
                 sex_col = "Sex")

print(example_data_checked)
```
If the input data is valid, you can proceed to the next steps.


## Step 3: Write the BIDS data

### Write the task files
Now, we use the write_task_tsv function to generate the BIDS-compliant directory and files.

This function performs the following actions:

1) Creates the output directory (if it doesn't already exist).
2) Writes a task-specific .tsv file with the behavioral data (response times, accuracy, etc.).

```{r}
# Output directory 
output_dir <- "example_bids"

# Write the task-related tsv files
write_task_tsv(data = example_data_checked, bids_dir = output_dir, 
               data_type = "beh", 
               filename_prefixes = c("sub-", "ses-", "task-"), 
               filename_variables = c("participant_id", "session", "reaction"="task"),
               path_prefixes = c("sub-", "ses-"), 
               path_variables = c("participant_id", "session"), 
               ignore_variables = c("age", "sex"))

```
You should find a directory now containing the desired files. Notice that we used a little trick here because we called out task "reaction" by providing a name in the vector. This allows defining variables that are invariant and not contained in the data. 
We also added the participant information to the "ignore_variables" because we want them in the participants.tsv but nowhere else. 

### Write the participants file

Next we will use the write_participants_tsv function to create a participants.tsv file that includes participant IDs and demographic variables.

```{r}
write_participants_tsv(data = example_data_checked, bids_dir = output_dir,
                       include_variables = c("participant_id", "age", "sex"))
```


## Step 4: Define and Write JSON Metadata

Next, we define the metadata for the task, which describe the behavioral variables recorded during the experiment (the task data contained in the tsv files).


The metadata is a nested list where each element describes a specific column in the dataset. The nested lists descrbibe keys belonging to that variable. 

```{r}
# Metadata for the example above with nested lists for each variable
meta_data <- list(
  participant_id = list(
    Description = "Unique participant identifier in the format sub-XXX.",
    Levels = list(
      "sub-1" = "Participant 1",
      "sub-2" = "Participant 2",
      "sub-3" = "Participant 3"
    )
  ),
  Sex = list(
    Description = "Self-reported Sex of the participant.",
    Levels = list(
      "m" = "Male",
      "f" = "Female",
      "o" = "Other"
    )
  ),
  age = list(
    Description = "Age of the participant in years.",
    Units = "years"
  ),
  session = list(
    Description = "Session identifier.",
    Levels = list(
      "ses-1" = "First session",
      "ses-2" = "Second session"
    )
  ),
  response_time = list(
    Description = "Response time of the participant in the task.",
    Units = "milliseconds"
  ),
  accuracy = list(
    Description = "Task accuracy (1 = correct, 0 = incorrect).",
    Levels = list(
      "0" = "Incorrect response",
      "1" = "Correct response"
    )
  )
)

# View the metadata
print(meta_data)
```

The JSON file that should be included in the BIDS structure can now easily be written wirth the function write_metadata(). 

```{r}
write_metadata(bids_dir = output_dir, task_name = "reaction", 
               data_type = "beh", 
               meta_data = meta_data)
```

A JSON file should have been added to the directory. 


## Step 5: Read and Inspect the Generated Files

Once the files are created, you can laod them again with the function read_bids().

```{r}
data_loaded <- read_bids(bids_dir = "example_bids", data_type = "beh", 
                         filename_prefixes =  c("sub-", "ses-", "task-"), 
                         filename_variables = c("participant_id", "session", "task"))
```
The output is a list with the participant data and the task data.

```{r}
data_loaded$participants
```

```{r}
data_loaded$task_data
```
We can see that it is identical to the data generated in Step 1. 
